
<!--
%\VignetteEngine{knitr::docco_linear}
%\VignetteIndexEntry{synlik_vignette}
-->
  
An Introduction to **mvn**
=======================================
  
```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(out.extra='style="display:block; margin: auto"', fig.align="center", tidy=FALSE)
```


Introduction
------------
  
The `mvn` R package provides computationally efficient tools related to the multivariate normal distribution. 
The tools are generally faster than those provided by other packages, thanks to the use of C++ code through the 
`Rcpp` package and of parallel computation through the `OpenMP` API.


Evaluating the multivariate normal density
----------------------------

Computing the density of multivariate normal is an essential step of MCMC or clustering algorithms, hence this
operations has to be as fast as possible. Here we compare the `dmvn` function with the equivalent function `dmvtnorm`, 
provided by the `mvtnorm` package. We start by evaluating the density of $10^2$ two-dimensional random vectors:
```{r dmvn}
library("microbenchmark")
library("mvtnorm")
library("mvn")

# Generating 
N <- 10000
d <- 50
mu <- 1:d
tmp <- matrix(rnorm(d^2), d, d)
mcov <- tcrossprod(tmp, tmp)
myChol <- chol(mcov)

microbenchmark(rmvn(N, mu, mcov, ncores = 4),
               rmvn(N, mu, mcov),
               rmvnorm(N, mu, mcov, method = "chol"),
               MASS::mvrnorm(N, mu, mcov))
```



References
----------------------------
  
  * Simon N Wood. Statistical inference for noisy nonlinear ecological dynamic systems. Nature, 466(7310):1102--1104, 2010.



















